# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
This file assumes that the data set is unzipped in the same location as the
source file (this file). Make sure to set your working directory.

First, let's read in the data file:
```{r}
activity_data <- read.csv('activity.csv', header=TRUE)
head(activity_data)
```

Now it's read in, we want to make sure that each of the variables (columns) is
in the appropriate form (i.e. data type). Let's take a look at what we have:

```{r}
summary(activity_data)
class(activity_data$steps)
class(activity_data$date)
class(activity_data$interval)
```

So we can see that steps and interval are integers, which is great. However,
the date variable would be more suitable as an actual date/time type for our
analysis later.

Let's convert it:

```{r}
activity_data$date <- as.Date(activity_data$date)
class(activity_data$date)
summary(activity_data)
```

Finally, let's convert it to a data table for some easy grouping later.
Note: you'll need the data.table package.

```{r}
library(data.table)
activity_dt <- data.table(activity_data)
```
## What is mean total number of steps taken per day?
Okay, now everything is loaded nicely, let's find the mean total number of steps
and print it:

```{r}
daily_steps <- activity_dt[,list(steps=sum(steps)),by=date]
head(daily_steps)
mean_steps <- mean(daily_steps$steps)
mean_steps
```

But, we can see that the NAs are causing problems. Let's ignore them for now:

```{r}
mean_steps <- mean(daily_steps$steps, na.rm=TRUE)
mean_steps
```

Now we can see that the mean is actually `r mean_steps`.
Similarly, let's look at a histogram of daily steps as well as the median.

```{r}
median_steps <- median(daily_steps$steps, na.rm=TRUE)
median_steps
hist(daily_steps$steps)
```

## What is the average daily activity pattern?
To look at the daily activity pattern, let's first group the average steps
by interval. Note that looking at the data, we have matching intervals on
each day, so we can just group these in our data table. Then we can plot the
time series as follows:

```{r}
interval_series <- activity_dt[,list(steps=mean(steps, na.rm=TRUE)),by=interval]
head(interval_series)
plot(steps ~ interval, data=interval_series, type ='l')
title(main='Average Steps By Interval')
```

Using our series, we can find the max average interval and plot a line to show
it. 

```{r}
max_interval <- interval_series[which.max(interval_series$steps),]
max_interval
plot(steps ~ interval, data=interval_series, type ='l')
title(main='Average Steps By Interval')
abline(v=max_interval$interval, col='red')
```

So, we see that the max interval is at `r max_interval$interval`. If we knew
that the 0 interval was midnight, we could find the corresponding time period.

## Imputing missing values

First, let's look at a summary of the original data set to look for NAs.
```{r}
summary(activity_data)
```

So it looks like date and interval are clean, but steps has 2304 NA values. We
can impute the missing data using the median for the corresponding interval.

```{r}
interval_medians <- activity_dt[,list(steps=median(steps, na.rm=TRUE)),by=interval]
cleaned_data <- activity_data
for (i in 1:nrow(cleaned_data)) {
        observation <- cleaned_data[i,]
        if (is.na(observation$steps)) {
                imputed <- interval_medians[interval_medians$interval == observation$interval,][1,]$steps
                cleaned_data[i,]$steps <- imputed
        }
}
summary(activity_data)
summary(cleaned_data)
```

So now we have imputed the data, let's repeat the mean, median, histogram work.

```{r}
clean_dt <- data.table(cleaned_data)
daily_steps_clean <- clean_dt[,list(steps=sum(steps)),by=date]
mean_steps_clean <- mean(daily_steps_clean$steps)
mean_steps_clean

median_steps_clean <- median(daily_steps_clean$steps)
median_steps_clean

hist(daily_steps_clean$steps)
```

So we can see that mean, median and distribution has changed rather dramatically
with the imputed data. The mean and median hae decreased, with an increase of
days with fewer steps and corresponding skew of the distribution to the left.

## Are there differences in activity patterns between weekdays and weekends?
To look for differences between weekdays and weekends, we need to first tag each
day as a weekday or weekend. We can then create our interval summary and plot
it.

```{r}
get_weekday <- function(date) {
        ifelse (weekdays(date) == 'Saturday' | weekdays(date) == 'Sunday', 
                'Weekend', 'Weekday')
}
cleaned_data$day <- as.factor(get_weekday(as.Date(cleaned_data$date)))
summary(cleaned_data)

cleaned_dt <- data.table(cleaned_data)
day_interval_series <- cleaned_dt[,list(steps=mean(steps, na.rm=TRUE)),by=list(interval,day)]
head(day_interval_series)

library(lattice)
xyplot(steps ~ interval | day, data=day_interval_series, type='l', layout=c(1,2),
       main="Steps by Interval for Weekends and Weekdays")
```

So from the plots, we can see that there is a higher peak period during weekdays, but more consistent activity over weekends.
